Computational methods in landscape ecology

### 2024-09-02 Sections/03-connectivity.qmd
Update references

# Connectivity

Connectivity describes how movement among resources patches is facilitated or hindered by landscape features such as types, amounts, and arrangement of LULC [@Taylor1993; @Taylor2006].
While structural connectivity focuses on physical and spatial relationships among resource patches, functional connectivity also includes the movement or flow of individuals and populations through the landscape [@Taylor2006].
Connectivity underpins ecosystem functionality and maintains biodiversity and populations, and plays a large role in many conservation actions [@CorreaAyram2016; @Hilty2020].
Thus, in the past multiple approaches how to measure connectivity were developed [@Simpkins2018; @Keeley2021].

@@

There are several landscape metrics that are traditionally used to measure landscape connectivity.
For example, the habitat amount within a buffer zone of focal patches, the nearest-neighbor distance, or the connectance index [@Moilanen2002; @Prugh2009; @Spanowicz2019].
However, these metrics are frequently criticised for being to simple and especially the link to functional connectivity is often ambiguous. 
Thus, more functionality needs to be included using additional ecological meaningful metrics, such as the core area [e.g., @Kupfer2012].
More recently there is also an increasing focus in within-patch connectivity using metrics such as the effective mesh size in addition to the usual measure between-patch connectivity [@Spanowicz2019].

@@

Resistance surfaces are one of the most popular approaches to model connectivity and developed from simple isolation-by-distance approaches. 
The approach uses environmental layers and biodiversity data or expert opinions to estimate the cost and of movement through landscapes [@Zeller2012].
The latest developments of resistance surfaces have recently been reviewed [@Dutta2022].
The resistance surface can be used to model movement based on different approaches such as leas-cost paths [@Adriaensen2003; @Etherington2016] or circuit theory [@McRae2008].
However, resistance surfaces are not free of critique including missing spatial-temporal variability or context dependency [@UnnithanKumar2022; @UnnithanKumar2022a].
Thus, recent developments include the combination with more processes-based modelling approaches [@Schumaker2018; @Bocedi2021; @UnnithanKumar2022b], modelling of dynamic landscape connectivity [@Zeller2020], inclusion of stochasticity and spatial context [@Ribeiro2017], using genetic optimization algorithms [@Peterman2018], or computational improvements [@Leonard2017].

@@

Graph theory describes a landscape as a set of nodes representing habitat patches and edges as potential links between the nodes.
Thus, graph theory is a powerful tool to model landscape connectivity, especially in a metapopulation context [@Urban2001].

>>>>

# Connectivity {#sec-connectivity}

Connectivity describes how movement among resources patches is facilitated or hindered by landscape features such as types, amounts, and arrangement of LULC [@Taylor1993ConnectivityVital; @Taylor2006LandscapeConnectivity].
While structural connectivity focuses on physical and spatial relationships among resource patches, functional connectivity also includes the movement or flow of individuals and populations through the landscape [@Taylor2006LandscapeConnectivity].
Connectivity underpins ecosystem functionality and maintains biodiversity and populations, and plays a large role in many conservation actions [@CorreaAyram2016HabitatConnectivity; @Hilty2020GuidelinesConserving].
Thus, in the past multiple approaches how to measure connectivity were developed [@Simpkins2018AssessingPerformance; @Keeley2021ConnectivityMetrics].

@@

There are several landscape metrics that are traditionally used to measure landscape connectivity.
For example, the habitat amount within a buffer zone of focal patches, the nearest-neighbor distance, or the connectance index [@Moilanen2002SimpleConnectivity; @Prugh2009EvaluationPatch; @Spanowicz2019MeasuringLandscape].
However, these metrics are frequently criticized for being to simple and especially the link to functional connectivity is often ambiguous. 
Thus, more functionality needs to be included using additional ecological meaningful metrics, such as the core area [e.g., @Kupfer2012LandscapeEcology].
More recently there is also an increasing focus in within-patch connectivity using metrics such as the effective mesh size in addition to the usual measure between-patch connectivity [@Spanowicz2019MeasuringLandscape].

@@

Resistance surfaces are one of the most popular approaches to model connectivity and developed from simple isolation-by-distance approaches. 
The approach uses environmental layers and biodiversity data or expert opinions to estimate the cost and of movement through landscapes [@Zeller2012EstimatingLandscape].
The latest developments of resistance surfaces have recently been reviewed [@Dutta2022OverviewComputational].
The resistance surface can be used to model movement based on different approaches such as leas-cost paths [@Adriaensen2003ApplicationLeastcost; @Etherington2016LeastcostModelling] or circuit theory [@McRae2008UsingCircuit].
However, resistance surfaces are not free of critique including missing spatial-temporal variability or context dependency [@UnnithanKumar2022ConnectivityModelling; @UnnithanKumar2022MovingLandscape].
Thus, recent developments include the combination with more processes-based modelling approaches [@Schumaker2018HexSimModeling; @Bocedi2021RangeShifterExtended; @UnnithanKumar2022PathwalkerNew], modelling of dynamic landscape connectivity [@Zeller2020UnderstandingImportance], inclusion of stochasticity and spatial context [@Ribeiro2017LandScapeCorridors], using genetic optimization algorithms [@Peterman2018ResistanceGAPackage], or computational improvements [@Leonard2017GlowSoftware].

@@

Graph theory describes a landscape as a set of nodes representing habitat patches and edges as potential links between the nodes.
Thus, graph theory is a powerful tool to model landscape connectivity, especially in a metapopulation context [@Urban2001LandscapeConnectivity].

### 2024-09-02 Sections/05-simulations.qmd
Update references

# Simulation models

Simulation models are a powerful tool to study complex adaptive ecological systems in controllable, reproducible, and replicable settings [@Pascual2005; @Petrovskii2012].
Thus, they can be seen as experimental systems that allow, unlike to the natural world, all imaginable manipulations in order to advance theoretical developments or test hypotheses [@Peck2004].
Due to the usual temporal and spatial scales, complex interactions and feedbacks, or scale mismatches between patterns and processes, simulation models are one of the major approaches in landscape ecology [@Schroder2006; @Synes2016].
In general, simulations models can be classified using two major divisions, namely as *i)* predictive or exploratory models, and *ii)* pattern- or process-based models [@Peck2004;@Synes2016].
Here, we are going to focus mainly on exploratory models, but include both pattern- and process-based models.

@@

Landscape simulators are typically used as null hypotheses, baselines, or scenarios because they allow to control certain aspects of the landscape [@Li2004a; @Wang2008] and are generally simpler than ecological simulations models [@Hess2020].
Landscape simulators can be classified into two major categories, namely pattern-based and process-based approaches [@Peer2013; @Langhammer2019].

Pattern-based approaches simulate landscapes without assuming any underlying abiotic or biotic processes [i.e., neutral landscape model\; @Gardner1987].
The earliest neutral models are based on percolation theory and assign LULC types randomly to cells in the landscape[@Gardner1987] or hierarchical models that consider different spatial scales while assigning cell values [@ONeill1992].
Landscapes characterized by continuous values can be simulated by fractal models, such as Brownian motion [@Palmer1992].
Borrowing from computer graphics, more recent neutral landscape models make use of spectral synthesis [e.g., Perlin noise\; @Etherington2022] or binary space partitioning [@Etherington2022a].
Neutral landscape models are also able to simulate landscapes dominated by anthropogenic activities using based on least-cost networks [@Etherington2024].
In order to ensure realistic neutral landscapes several approaches exist that use comparisons with real landscapes [@Inkoom2017] or are based on target values optimization [@VanStrien2016; @Justeau-Allaire2022].
<!-- In order to make multiple neutral landscape model algorithms accessible, several software libraries are available in prominent programming languages including Python [NLMpy\; @Etherington2015], R [ c], Julia (https://github.com/EcoJulia/NeutralLandscapes.jl), or GUI software (GradientLand, QRULE) -->

Contrastingly, processes-based approaches explicitly include abiotic or biotic pattern-forming processes [i.e., landscape generators\; @Langhammer2019; @Salecker2019].
Several earlier landscape generators are based on cellular automata models and are able to simulate urban growth [@Clarke1998] or deforestation [@Soares-Filho2002].
More recent landscape generators allow to simulate patchy landscapes based on Gibbs processes [@Gaucherel2006], deforestation based on road and agricultural fields access [@Peer2013], agricultural areas dominated by smallholders [@Salecker2019], or vegetation surround watering points in semi-arid savanna rangelands [combining pattern- and process-bases approaches\; @Hess2020].

@@

Individual-based models (or agent-based models) simulate discrete entities that are described by attributes and behaviors and patterns emerge from bottom-up interactions of individuals with each other and their environment [@DeAngelis2014].
In landscape ecology individual-based models are increasingly used to model especially social-ecological systems [e.g, @Carauta2018; @Dislich2018; @Gonzalez-Redin2019], but also disturbances [e.g., @Seidl2017a; @Spies2017; @Kamimura2019] or connectivity [e.g., @Day2020].
In order to facilitate future model developments, recent progress includes adaptable modeling frameworks [@Schumaker2018] or reusable building block [@Berger2024].
Furthermore, hybrid approaches that couple or even integrated different model types with individual-based models could benefit from rich development history or facilitate corresponding strengths [@OSullivan2016], e.g., linking an individual-based and ecosystem model [@Polhill2011].

@@

Digital Twins include three connected elements, namely a physical object or system, its digital representation, and a data exchange between the former two [@Jones2020; @DeKoning2023] and there is an increasing interest and use of them in both the industry and academia [@Jones2020].
The need for timely and evidence-based decision making in combination with increasing data availability makes them a powerful tool for ecology [@DeKoning2023].
Digital twins are increasingly applied in agricultural landscapes research, including topics such as crop and land cultivation, livestock farming, controlled environment farming, or fertilization management [@Cesco2023; @Purcell2023].
Further applications include planning of rural ecological landscapes [@Tan2024], or exploration of relationships between urban expansion and vegetation coverage [@Zhao2022].
The use of general data processing, statistical and mechanistic models, or artificial intelligence allows them to constantly update the digital representation and analyze, visualize, or predict the physical counterpart.
Thus, Digital Twins can provide real-time state of nature measurements, early detection of conservation trends, relationships of ecosystem trends and environmental conditions, impact assessments of interventions, or identify uncertainties and information gaps [@DeKoning2023].

>>>>

# Simulation models {#sec-simulation}

Simulation models are a powerful tool to study complex adaptive ecological systems in controllable, reproducible, and replicable settings [@Pascual2005ComputationalEcology; @Petrovskii2012ComputationalEcology].
Thus, they can be seen as experimental systems that allow, unlike to the natural world, all imaginable manipulations in order to advance theoretical developments or test hypotheses [@Peck2004SimulationExperiment].
Due to the usual temporal and spatial scales, complex interactions and feedbacks, or scale mismatches between patterns and processes, simulation models are one of the major approaches in landscape ecology [@Schroder2006AnalysisPattern; @Synes2016EmergingOpportunities].
In general, simulations models can be classified using two major divisions, namely as *i)* predictive or exploratory models, and *ii)* pattern- or process-based models [@Peck2004SimulationExperiment; @Synes2016EmergingOpportunities].
Here, we are going to focus mainly on exploratory models, but include both pattern- and process-based models.

@@

Landscape simulators are typically used as null hypotheses, baselines, or scenarios because they allow to control certain aspects of the landscape [@Li2004EvaluatingEffectiveness; @Wang2008NeutralLandscapes] and are generally simpler than ecological simulations models [@Hess2020PioLaGPiosphere].
Landscape simulators can be classified into two major categories, namely pattern-based and process-based approaches [@Peer2013SimpleProcessbased; @Langhammer2019AgriculturalLandscape].

Pattern-based approaches simulate landscapes without assuming any underlying abiotic or biotic processes [i.e., neutral landscape model\; @Gardner1987NeutralModels].
The earliest neutral models are based on percolation theory and assign LULC types randomly to cells in the landscape[@Gardner1987NeutralModels] or hierarchical models that consider different spatial scales while assigning cell values [@ONeill1992HierarchicalNeutral].
Landscapes characterized by continuous values can be simulated by fractal models, such as Brownian motion [@Palmer1992CoexistenceSpecies].
Borrowing from computer graphics, more recent neutral landscape models make use of spectral synthesis [e.g., Perlin noise\; @Etherington2022PerlinNoise] or binary space partitioning [@Etherington2022BinarySpace].
Neutral landscape models are also able to simulate landscapes dominated by anthropogenic activities using based on least-cost networks [@Etherington2024LeastcostNetwork].
In order to ensure realistic neutral landscapes several approaches exist that use comparisons with real landscapes [@Inkoom2017DesigningNeutral] or are based on target values optimization [@VanStrien2016ImprovedNeutral; @Justeau-Allaire2022FragmentedLandscape].
<!-- In order to make multiple neutral landscape model algorithms accessible, several software libraries are available in prominent programming languages including Python [NLMpy\; @Etherington2015NLMpyPython], R [NLMR; @Sciaini2018NLMRLandscapetools], Julia (https://github.com/EcoJulia/NeutralLandscapes.jl), or GUI software (GradientLand, QRULE) -->

Contrastingly, processes-based approaches explicitly include abiotic or biotic pattern-forming processes [i.e., landscape generators\; @Langhammer2019AgriculturalLandscape; @Salecker2019EFForTSLGrafLandscape].
Several earlier landscape generators are based on cellular automata models and are able to simulate urban growth [@Clarke1998LoosecouplingCellular] or deforestation [@Soares-Filho2002DinamicaStochastic].
More recent landscape generators allow to simulate patchy landscapes based on Gibbs processes [@Gaucherel2006NeutralModels], deforestation based on road and agricultural fields access [@Peer2013SimpleProcessbased], agricultural areas dominated by smallholders [@Salecker2019EFForTSLGrafLandscape], or vegetation surround watering points in semi-arid savanna rangelands [combining pattern- and process-bases approaches\; @Hess2020PioLaGPiosphere].

@@

Individual-based models (or agent-based models) simulate discrete entities that are described by attributes and behaviors and patterns emerge from bottom-up interactions of individuals with each other and their environment [@DeAngelis2014IndividualbasedModels].
In landscape ecology individual-based models are increasingly used to model especially social-ecological systems [e.g, @Carauta2018CanPreferential; @Dislich2018LanduseChange; @Gonzalez-Redin2019ExploringSustainable], but also disturbances [e.g., @Seidl2017ClimateChange; @Spies2017UsingAgentbased; @Kamimura2019AgentbasedModelling] or connectivity [e.g., @Day2020IndividualbasedModeling].
In order to facilitate future model developments, recent progress includes adaptable modeling frameworks [@Schumaker2018HexSimModeling] or reusable building block [@Berger2024ReusableBuilding].
Furthermore, hybrid approaches that couple or even integrated different model types with individual-based models could benefit from rich development history or facilitate corresponding strengths [@OSullivan2016StrategicDirections], e.g., linking an individual-based and ecosystem model [@Polhill2011AgentbasedModelling].

@@

Digital Twins include three connected elements, namely a physical object or system, its digital representation, and a data exchange between the former two [@Jones2020CharacterisingDigital; @DeKoning2023DigitalTwins] and there is an increasing interest and use of them in both the industry and academia [@Jones2020CharacterisingDigital].
The need for timely and evidence-based decision making in combination with increasing data availability makes them a powerful tool for ecology [@DeKoning2023DigitalTwins].
Digital twins are increasingly applied in agricultural landscapes research, including topics such as crop and land cultivation, livestock farming, controlled environment farming, or fertilization management [@Cesco2023SmartAgriculture; @Purcell2023DigitalTwins].
Further applications include planning of rural ecological landscapes [@Tan2024DigitalTwin], or exploration of relationships between urban expansion and vegetation coverage [@Zhao2022ApplyingDigital].
The use of general data processing, statistical and mechanistic models, or artificial intelligence allows them to constantly update the digital representation and analyze, visualize, or predict the physical counterpart.
Thus, Digital Twins can provide real-time state of nature measurements, early detection of conservation trends, relationships of ecosystem trends and environmental conditions, impact assessments of interventions, or identify uncertainties and information gaps [@DeKoning2023DigitalTwins].

### 2024-09-02 Sections/08-software.qmd
Update Bib and manuscript
Process to connectivity section
Update references

<!-- Person responsible: Hesselbarth, Nowosad, Bosch, Jung, Schouten  -->

@@

<!-- "The future applications of landscape pattern metrics will almost certainly depend on the software and analysis tools available" -- @costanzaDescribingAnalyzingLandscape2019 -->

"Open science" involves full transparency and data access at all stages of the research processes and the integrative and multi-disciplinary field of ecology can massively benefit from this concept [@Hampton2015].
One key aspect of achieving open science is to use open source software, which ensures transparency and consequently increasing the (public) trust into the analyses [@Powers2019].
Related to this, script-based workflows have the advantage over workflows involving graphical user interfaces that the former are shareable and reproducible [@Hampton2015; @Powers2019].
Thus, we will focus the following section on script-based open-source software suitable for landscape ecology.
Given the increasing availability of data and complexity of methods, future progress within the field of landscape ecology will depend on the software and analysis tools available [@Costanza2019]

<!-- R: Hesselbarth -->
The open-source and script-based R programming languages [@RCoreTeam2024] was first released in 1995 [@Smith2016] with a focus on statistical computing and data visualization.
However, due to its extensions (also called packages), the language is popular in many academic fields including ecology [@Lai2019] and it is suitable for spatial data analysis or landscape ecology [e.g., @Bivand2020; @Hesselbarth2021].

<!-- 
landscapemetrics
geodiv
vectormetrics (also: VecLI -- they mention open source in their paper a few times, but the software seems to be closed source -- https://urbancomp.net/archives/vecliv3beta)
bespatial
LandComp
motif
?GLCMTextures
?rasterdiv
?terra 
-->

<!-- Python: Bosch -->

<!--Julia: Schouten -->

<!--Others: -->

<!-- 
GuidosToolbox
?GRASS GIS
?QGIS: LecoS plugin 
-->

>>>>

<!-- Person responsible: Hesselbarth, Nowosad, Bosch, Jung  -->

@@

"Open science" involves full transparency and data access at all stages of the research processes and integrative and multi-disciplinary fields such as ecology can massively benefit from this concept [@Hampton2015TaoOpen].
Using open source software is one key aspect of open science because it ensures transparency and consequently increases trust in the analyses [@Powers2019OpenScience].
Given the increasing data availability and method complexity, future progress within the research field will highly depend on available software and analysis tools [@Costanza2019DescribingAnalyzing].
In this context, script-based workflows have an advantage over workflows involving graphical user interfaces because the former are more shareable and reproducible [@Hampton2015TaoOpen; @Powers2019OpenScience].
Thus, we will focus the following section on script-based open-source software suitable for landscape ecology.

<!-- R: Hesselbarth -->
The open-source, script-based R programming languages [@RCoreTeam2024LanguageEnvironment] was first released in 1995 [@Smith201616Years] with a focus on statistical computing and data visualization.
However, due to its extensions (also called packages), the language is popular in many academic fields, including ecology [@Lai2019EvaluatingPopularity; @Lortie2020ChecklistChoosing], and it is suitable for spatial data analysis and landscape ecology [e.g., @Bivand2020ProgressEcosystem; @Hesselbarth2021OpensourceTools].
Fundamental packages for spatial data analysis include *sf* for simple-features vector data [@Pebesma2018SimpleFeatures], *terra* for raster data [@Hijmans2024TerraSpatial], *stars* for spatiotemporal arrays, raster, and vector data cubes [@Pebesma2023SpatialData], and *spatstat* for point pattern data [@Baddeley2015SpatialPoint].
Several packages allow to create maps, such as *tmap* [@Tennekes2018TmapThematic] or *ggplot2* [@Wickham2016Ggplot2Elegant]. 
For a comprehensive overview of packages related to geographical spatial data, please see the CRAN Task View [@Bivand2024CRANTask].
R packages developed specifically for spatial pattern analysis include, besides others, *landscapemetrics* [@Hesselbarth2019LandscapemetricsOpen], *vectormetrics* [still being developed; @Matuszek2024VectormetricsLandscape], *motif* [@Nowosad2021MotifOpensource], *rasterdiv* [@Rocchini2021RasterdivInformation], or *geodiv* [@Smith2021GeodivPackage]. 
<!-- BESSpatial: https://github.com/Nowosad/bespatial/ -->
Packages with a focus on scale include e.g., *multilandr* [@Huais2024MultilandrPackage] or *grainchanger* [@Graham2024GrainchangerMovingwindow].
Connectivity analysis can be conducted using packages including *lconnect* [@Mestre2023LconnectPackage], or *grainscape* [@Chubaty2020ToolboxGrainscape].
Furthermore, neutral landscapes can be simulated using *NLMR* [@Sciaini2018NLMRLandscapetools], and the *graph4lg* [@Savary2021Graph4lgPackage] provides functions for landscape genetics.

<!-- Python: Bosch, Jung? -->
